---
title: "PHASE_THREE_STATISTICAL_ANALYSIS_AND_GRAPHS"
author: "Chris Low"
date: "01/09/2021"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr) # data visualization
library(tidyverse) # data wrangling and visualization
library(ggplot2) # data visualization
library(performance) # check model outputs
library(lme4) # analysis for logistic regressions and mixed-effects modelling
library(corrplot) # creating a correlation matrix
library(sjPlot) # to visualizing mixed-effects models
library(effects) # to visualizing mixed-effects models
library(report) # mainly for an "report" function
library(knitr) # beautifying tables
library(MASS) # model selection using drop terms
library(pscl) # zero-inflation regression models
library(MuMIn) # generating marginal and conditional R2 values
library(DHARMa) # model criticism plots
library(effects) # extracting model outputs
library(jtools) # 
library(webshot) # extract SJplots - for tables
library(StatisticalModels) # package by Tim Newbold fro assessing statistical 
library(gridExtra) # sort out gridding for plots
library(ggthemes) # make nice themed plots
library(RColorBrewer) # make colour palletes


```

# Create plotting themes
```{r plotting theme, echo=TRUE}
theme_Publication <- function(base_size=14, base_family="helvetica") {
      library(grid)
      library(ggthemes)
      (theme_foundation(base_size=base_size, base_family=base_family)
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               legend.key.size= unit(0.2, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}


```


## Loading in final dataset in r

```{r load data, echo=TRUE}

# Just save priority EDGE lineage list and PD_mammal_merged list as essential for analysis: mammals_pd_merged, EDGE_priority_lineages, ePD_priority_lineages (final datasets so far!!)
load("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Data and analysis/EDGE_families_and_scores/EDGE_LINEAGES_FOR_ANALYSIS.RData")

# load dataout datasets with both Upham and Gumbs datasets
load("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Data and analysis/Mammal phylogenies/ePD_scores_ALL.RData")

```

# Summary statistics of entirly threatened lineages
```{r threatened lienages}

# subset famileis which entire population are threatened with extinction
families_threatened <- mammals_pd_merged[which(mammals_pd_merged$assessed_threat_prop == 1),]

# select important columns
families_threatened <- families_threatened[,c(1:3,7,9,14)]

# subset families that have all been assessed
families_threatened <- families_threatened[which(families_threatened$unassessed_prop == 0),]

# sum ePD loss of these families (384 MY)
monotypic_spp <- subset(families_threatened, families_threatened$richness ==1)

# sum ePD of monotypic species (174 MY)
sum(monotypic_spp$Median.exPD)

# caculate conservation staus for mammals in 2008 to see what uplisting there have been since then
uplistming_sub <- RL_change_df[(RL_change_df$Family %in% families_threatened$Family),]

# three families weren't in intial 2008 RL assessment - don't include these as I think they were already highlighted for importance such as Lipotidae (river dolphin)
list(setdiff(families_threatened$Family,uplistming_sub$Family))

# 10 families uplisted since 2008 - 40% of families have had uplistings since 2008
sum(uplistming_sub$prop_uplisted > 0)

# 5 families have have become fully threatened since 2008 (20%)
sum(uplistming_sub$prop_uplisted == 1)

sum(families_threatened$Median.exPD)

```

## Testing the difference between

You can also embed plots, for example:

```{r testing phylogeny datasets, echo=FALSE}

# Spearmans's rank correaltion between imputted an non-imputed datasets
cor.test(ex_PD_full$Median.exPD, ex_PD_full$Median.ePD.UPHAM, method = "spearman")

# subset variables for analysis
ePD_spp_rich_df <- mammals_pd_merged[,c(3,14,39,49)]

# Spearman's rank correlation between species richness and ePD (rs = 0.67; p < 0.001; N = 160)
cor.test(ePD_spp_rich_df$richness, ePD_spp_rich_df$Median.exPD, method = "spearman") 

# plot for species richness and ePD loss
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Imperilled PD/ePDloss_richness_FINAL.tiff", units = "in", width = 5, height = 4, res = 300)

ggplot(ePD_spp_rich_df, aes(x=richness, y=Median.exPD)) + geom_point(size=2, shape=20) +
  geom_smooth(method=lm, se = FALSE, color = "red") +
  scale_y_log10() +
  scale_x_log10() +
  theme_bw(base_size = 16) +
  labs(y="(Log) Median ePD Loss", x= "(Log) Richness")

dev.off()

# Spearman's rank correaltion for EDGE score (rs = -0.61; p < 0.001; N = 160)
cor.test(ePD_spp_rich_df$richness, ePD_spp_rich_df$median_EDGE, method = "spearman")

# scatterplot for EDGE score and spp richness
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Imperilled PD/EDGE_richness_FINAL.tiff", units = "in", width = 5, height = 4, res = 300)

ggplot(ePD_spp_rich_df, aes(x=richness, y=median_EDGE)) + geom_point(size=2, shape=20) +
  geom_smooth(method=lm, se = FALSE, color = "red") +
  scale_y_log10() +
  scale_x_log10() +
  theme_bw(base_size = 16) +
  labs(y="(Log) Median EDGE Score", x= "(Log) Richness")

dev.off()

# Spearman's rank correaltion for ED scores (rs = -0.73; p < 0.001; N = 160)
cor.test(ePD_spp_rich_df$richness, ePD_spp_rich_df$median_ED, method = "spearman")

# scatterplot for ED and species richness
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Imperilled PD/ED_richness_FINAL.tiff", units = "in", width = 5, height = 4, res = 300)

ggplot(ePD_spp_rich_df, aes(x=richness, y=median_ED)) + geom_point(size=2, shape=20) +
  geom_smooth(method=lm, se = FALSE, color = "red") +
  scale_y_log10() +
  scale_x_log10() +
  theme_bw(base_size = 16) +
  labs(y="(Log) Median ED Score", x= "(Log) Richness")

dev.off()

```
## Question (i) How do intrinsic and extrinsic threats predict ED?

```{r intrinsic and extrinsic threats, echo=TRUE}

# subset explanatory variables needed for this question
ePD_response_df <- mammals_pd_merged[,c(1,2,27,29,31,33,35,37:39,48:49)]

# Calculate number of occurences which are greater than 0 to sum threat data
ePD_response_df$threat_sum <- apply(ePD_response_df[,c("prop_1_1","prop_2_1","prop_2_3","prop_5_1","prop_5_3","prop_8_1")], 1, function(x) length(which(x > 0)))

# create cum sum of threats as a proportion
ePD_response_df <- ePD_response_df %>% 
  group_by(Family) %>% 
  mutate(threat_sum_prop = threat_sum/6)

ePD_response_df$threat_sum <- as.numeric(ePD_response_df$threat_sum)

# visulising distributions of data
# distribution of response variable is poission
par(mfrow=c(3:4))
for(i in 3:14) {
    hist(ePD_response_df[,i], main=names(ePD_response_df)[i])
}

# check colinearity of data
# creating a correlation matrix
correlations <- cor(ePD_response_df[,3:14])
corrplot(correlations, method="circle")

# Remove NA's
ePD_response_df <- ePD_response_df[complete.cases(ePD_response_df),]

# run a glm- MAXIMAL MODEL
ED_glmm_full <- lmer(log(median_ED + 1) ~ log(bodysize_mean + 1) + log(Gen_length_days +1) + prop_1_1 + prop_2_1 + prop_2_3 + prop_5_1 + prop_5_3 + prop_8_1 + threat_sum + (1| Order),
              na.action = "na.fail",
              data = ePD_response_df) 

summary(ED_glmm_full)

tab_model(ED_glmm_full)

# find out explanatory power
StatisticalModels::R2GLMER(ED_glmm_full)
MuMIn::r.squaredGLMM(ED_glmm_full)


# check model performance - assumptions not violated - no they are not xx
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_crit.tiff", units = "in", width = 8, height = 10, res = 300)

performance::check_model(ED_glmm_full)

dev.off()

# update the new model
ED_glmm_new <- lmer(log(median_ED + 1) ~ log(bodysize_mean + 1) + log(Gen_length_days +1) + prop_1_1 + prop_2_1 + prop_5_1 + prop_5_3 + prop_8_1 + threat_sum_prop + (1| Order),
              na.action = "na.fail",
              data = ePD_response_df) 

tab_model(ED_glmm_full)

# find out explanatory power
StatisticalModels::R2GLMER(ED_glmm_full)

# Using anova to compare model fits (chi sqr test) - keep full model as it has the highest explanatory power - keep ful model
anova(ED_glmm_new, ED_glmm_full, test = "chi")

# create null model
ED_null <- lmer(log(median_ED +1) ~ 1 + (1 | Order), REML = FALSE, data = ePD_response_df)

# compare full and null model
anova(ED_glmm_full, ED_null, test = "chi")

### create tabular format or results
tab_model(file = "ED_glmm_table.html",
  ED_glmm_full,
  CSS = list(
    css.depvarhead = 'color: black;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: black;'
  )
)

webshot("ED_glmm_table.html", "ED_glmm_table.png")

## extract effects of significant effects in the model and predcit them
# Plotting the response of ED 
# pull out the significant effects of the model
ED_effects_2_1 <- effects::effect(term = c("prop_2_3"), mod = ED_glmm_full) %>%
  # turn it into a dataframe
  as.data.frame()

ED_effects_2_1$Predictor <- "Threat 2.3"
names(ED_effects_2_1)[1] <- "proportion"

ED_effects_5_3 <- effects::effect(term = c("prop_5_3"), mod = ED_glmm_full) %>%
  # turn it into a dataframe
  as.data.frame()

ED_effects_5_3$Predictor <- "Threat 5.3"
names(ED_effects_5_3)[1] <- "proportion"


ED_effects_8_1 <- effects::effect(term = c("prop_8_1"), mod = ED_glmm_full) %>%
  # turn it into a dataframe
  as.data.frame()

ED_effects_8_1$Predictor <- "Threat 8.1"
names(ED_effects_8_1)[1] <- "proportion"


ED_effects_threat_sum <- effects::effect(term = c("threat_sum"), mod = ED_glmm_full) %>%
  # turn it into a dataframe
  as.data.frame()

ED_effects_threat_sum$Predictor <- "Threat sum"
names(ED_effects_threat_sum)[1] <- "proportion"

# merge dataframes in long format
ED_glmmm_df <- rbind(ED_effects_2_1, ED_effects_5_3, ED_effects_8_1)

# remember to back-transform variables!!
ED_glmmm_df$fit_log <- exp(ED_glmmm_df$fit)
ED_glmmm_df$upper_log <- exp(ED_glmmm_df$upper)
ED_glmmm_df$lower_log <- exp(ED_glmmm_df$lower)

tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_predict.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot
p <- ggplot(ED_glmmm_df) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit_log, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower_log, ymax = upper_log, group = Predictor, fill = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Proportion of family" ,y = "Median ED")

grid.arrange((p + scale_colour_brewer(palette = "Set1") + theme_Publication()),nrow=1)

dev.off()

# run for the the acummulative sum for threats
ED_effects_threat_sum$log_fit <- exp(ED_effects_threat_sum$fit)
ED_effects_threat_sum$log_upper <- exp(ED_effects_threat_sum$upper)
ED_effects_threat_sum$log_lower <- exp(ED_effects_threat_sum$lower)


tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_threat_sum_predict.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot
p <- ggplot(ED_effects_threat_sum) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = log_fit, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = log_lower, ymax = log_upper, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Number of threats" ,y = "Median ED")

grid.arrange((p + scale_colour_brewer(palette = "Set1") + theme_Publication()),nrow=1)

dev.off()

```

# same models but for EDGE scores
```{r EDGE GLMM}
# Check if intrinsic and extrintic threats respond to EDGE score
# run a glmm with EDGE score as the response variable- MAXIMAL MODEL
EDGE_glmm_full <- lmer(log(median_EDGE + 1) ~ log(bodysize_mean + 1) + log(Gen_length_days +1) + prop_1_1 + prop_2_1 + prop_2_3 + prop_5_1 + prop_5_3 + prop_8_1 + threat_sum + (1|Order),
              na.action = "na.fail",
              data = ePD_response_df) 

# get model summary
summary(EDGE_glmm_full)
tab_model(EDGE_glmm_full)

# find out explanatory power
StatisticalModels::R2GLMER(EDGE_glmm_full)

# check model performance - assumptions not violated - no they are not xx
# check model performance - assumptions not violated - no they are not xx
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/EDGE_glmm_crit.tiff", units = "in", width = 8, height = 10, res = 300)

performance::check_model(EDGE_glmm_full)

dev.off()

# backwards stepwise model selection

EDGE_glmm_new <- lmer(log(median_EDGE + 1) ~ log(bodysize_mean + 1) + prop_1_1 + prop_2_1 + prop_2_3 + prop_5_1 + prop_5_3 + prop_8_1 + threat_sum + (1|Order),
              na.action = "na.fail",
              data = ePD_response_df)

# get model summary
summary(EDGE_glmm_new)

# find out explanatory power
StatisticalModels::R2GLMER(EDGE_glmm_new)

# Compare full and updated model- USE FULL MODEL
anova(EDGE_glmm_full, EDGE_glmm_new)

# look at summary table
tab_model(EDGE_glmm_new)

# Create null model
EDGE_glmm_null <- lmer(log(median_EDGE + 1) ~ 1 + (1|Order), 
                       na.action = "na.fail",
                       data = ePD_response_df) 

# compare full model to the null- full model is signifcant
anova(EDGE_glmm_full, EDGE_glmm_null)

### plotting outputs for EDGE scores and extrinsic threats
### create tabular format or results
tab_model(file = "EDGE_glmm_table.html", ED_glmm_full,
  EDGE_glmm_full,
  CSS = list(
    css.depvarhead = 'color: black;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: black;'
  )
)

webshot("EDGE_glmm_table.html", "EDGE_glmm_table.png")

## extract effects of significant effects in the model and predcit them
# Plotting the response of EDGE
# pull out the significant effects of the model
EDGE_effects_5_3 <- effects::effect(term = c("prop_5_3"), mod = EDGE_glmm_full) %>%
  # turn it into a dataframe
  as.data.frame()

EDGE_effects_5_3$Predictor <- "Threat 5.3"
names(EDGE_effects_5_3)[1] <- "proportion"


EDGE_effects_8_1 <- effects::effect(term = c("prop_8_1"), mod = EDGE_glmm_full) %>%
  # turn it into a dataframe
  as.data.frame()

EDGE_effects_8_1$Predictor <- "Threat 8.1"
names(EDGE_effects_8_1)[1] <- "proportion"


EDGE_effects_threat_sum <- effects::effect(term = c("threat_sum"), mod = EDGE_glmm_full) %>%
  # turn it into a dataframe
  as.data.frame()

EDGE_effects_threat_sum$Predictor <- "Threat sum"
names(EDGE_effects_threat_sum)[1] <- "proportion"

# merge dataframes in long format
EDGE_glmmm_df <- rbind(EDGE_effects_5_3, EDGE_effects_8_1)

# remember to back-transform variables!!
EDGE_glmmm_df$fit_log <- exp(EDGE_glmmm_df$fit)
EDGE_glmmm_df$upper_log <- exp(EDGE_glmmm_df$upper)
EDGE_glmmm_df$lower_log <- exp(EDGE_glmmm_df$lower)

tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/EDGE_glmm_predict.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot
p <- ggplot(EDGE_glmmm_df) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit_log, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower_log, ymax = upper_log, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Proportion of family" ,y = "Median EDGE")

grid.arrange((p + scale_colour_brewer(palette = "Set1") + theme_Publication()),nrow=1)

dev.off()

#setup ggplot for threat sum
# remember to back-transform variables!!
EDGE_effects_threat_sum$fit_log <- exp(EDGE_effects_threat_sum$fit)
EDGE_effects_threat_sum$upper_log <- exp(EDGE_effects_threat_sum$upper)
EDGE_effects_threat_sum$lower_log <- exp(EDGE_effects_threat_sum$lower)

tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/EDGE_glmm_threat_sum_predict.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot
p <- ggplot(EDGE_effects_threat_sum) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit_log, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower_log, ymax = upper_log, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Number of threats" ,y = "Median EDGE")

grid.arrange((p + scale_colour_brewer(palette = "Set1") + theme_Publication()),nrow=1)

dev.off()

# get combination of model tables to decide which one to use
tab_model(file = "EDGE_ED_glmm_all_table.doc",
  EDGE_glmm_full, ED_glmm_full,
  string.est = "Estiamte",
  digits = 4,
  CSS = list(
    css.depvarhead = 'color: black;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: black;'
  )
)

webshot("EDGE_ED_glmm_table.html", "EDGE_ED_glmm_table.png")


```

## Question (ii) How does the inteaction between EDGE/ED score with proportion of threatened species predict conservation status?


```{r conservation staus, echo=TRUE}

# function to test for overdispersion
overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

# subset data to only contain population probabilities
con_status_interacts_df <- mammals_pd_merged[,c(1,2,9,11,21,23,25,39,49)]

# Remove all NA's from columns
con_status_interacts_df <- con_status_interacts_df[complete.cases(con_status_interacts_df),]

# add column to add observational level occurences
con_status_interacts_df$observation <- 1:nrow(con_status_interacts_df)

con_status_interacts_df$observation <- as.numeric(con_status_interacts_df$observation)

# visulising distibutions of data
par(mfrow=c(4,2))
for(i in 3:9) {
    hist(con_status_interacts_df[,i], main=names(con_status_interacts_df)[i])
}

# creating a correlation matrix
correlations <- cor(con_status_interacts_df[,3:9])
corrplot(correlations, method="circle")

# double check strucutre of dataset
str(con_status_interacts_df)

# binomial glmm in which EDGE or ED scores is a responds to proportion of threatened species interacts with threats
# perform a glm fit

con_status_uplisted_glmm_ED <- glmer(prop_uplisted ~ assessed_threat_prop*median_ED + (1 | Order), data = con_status_interacts_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

# test for overdispersion = not present
overdisp_fun(con_status_uplisted_glmm_ED)
check_overdispersion(con_status_uplisted_glmm_ED)

# get summary of model
summary(con_status_uplisted_glmm_ED)

# check model
check_model(con_status_uplisted_glmm_ED)

tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/con_status_uplisted_glmm_ED_crit.tiff", units = "in", width = 8, height = 10, res = 300)

performance::check_model(con_status_uplisted_glmm_ED)

dev.off()

# tabular format of results
tab_model(con_status_uplisted_glmm_ED, show.est = TRUE, string.est = "Estimate", transform = NULL, digits = 4)

# get conditional R2 values
StatisticalModels::R2GLMER(con_status_uplisted_glmm_ED)
MuMIn::r.squaredGLMM(con_status_uplisted_glmm_ED)

# create null model
con_status_uplisted_glmm_null <- glmer(prop_uplisted ~ 1 + (1 | Order), data = con_status_interacts_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

summary(con_status_uplisted_glmm_null)

# compare full and null model fit- keep full model
anova(con_status_uplisted_glmm_ED, con_status_uplisted_glmm_null)

### plotting outputs for conservation status: ED scores and threatened species
### create tabular format or results
tab_model(file = "con_status_uplisted_glmm_ED.html",
  con_status_uplisted_glmm_ED,
  show.est = TRUE,
  string.est = "Estiamte",
  transform = NULL,
  digits = 4,
  CSS = list(
    css.depvarhead = 'color: black;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: black;'
  )
)

webshot("con_status_uplisted_glmm_ED.html", "con_status_uplisted_glmm_ED.png")

# Predict conservation up-listings for ED:threatened species interaction
# pull out the significant effects of the model
ED_uplisting_df_threat <- effects::effect(term = c("assessed_threat_prop"), mod = con_status_uplisted_glmm_ED) %>%
  # turn it into a dataframe
  as.data.frame()

ED_uplisting_df_threat$Predictor <- "Threatened species"
names(ED_uplisting_df_threat)[1] <- "proportion"

# do same for ED
ED_uplisting_df_ED <- effects::effect(term = c("median_ED"), mod = con_status_uplisted_glmm_ED) %>%
  # turn it into a dataframe
  as.data.frame()

ED_uplisting_df_ED$Predictor <- "Median ED"
names(ED_uplisting_df_ED)[1] <- "proportion"

tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_prop__fam_threat.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot to plot proportion of threatened species and conservation up-listings
p <- ggplot(ED_uplisting_df_threat) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower, ymax = upper, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Proportion of family threatened" ,y = "Proportion of Red List up-listings")

# change colour pallete to orange
grid.arrange((p + scale_colour_brewer(palette = "Spectral", direction = -1) + theme_Publication()),nrow=1)

dev.off()

tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_prop_ED_uplisting.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot to predict median EDGE score and proportion of con uplistings
p <- ggplot(ED_uplisting_df_ED) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower, ymax = upper, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Median ED" ,y = "")

# change colour pallete to orange
grid.arrange((p + scale_colour_brewer(palette = "Set1") + theme_Publication()),nrow=1)

dev.off()
```


# running conservation up-listing GLMM for EDGE scores
```{R GLMM EDGE scores}
# Just fitting EDGE score as a predictor
con_status_uplisted_glmm_EDGE <- glmer(prop_uplisted ~ median_EDGE + (1 | Order), data = con_status_interacts_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

# get summary of results
summary(con_status_uplisted_glmm_EDGE)

# get model tab
tab_model(con_status_uplisted_glmm_EDGE, show.est = TRUE, string.est = "Estimate", transform = NULL, digits = 4)

# get explanatory power of model 
StatisticalModels::R2GLMER(con_status_uplisted_glmm_EDGE)
MuMIn::r.squaredGLMM(con_status_uplisted_glmm_ED)

# check for model performance
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/con_status_uplisted_glmm_EDGE_crit.tiff", units = "in", width = 8, height = 10, res = 300)

performance::check_model(con_status_uplisted_glmm_EDGE)

dev.off()

# I think there is overdispersion still is overdispersion however is better than preious model
overdisp_fun(con_status_uplisted_glmm_EDGE)
check_overdispersion(con_status_uplisted_glmm_EDGE)


# compare with null model
con_status_uplisted_glmm_EDGE_null <- glmer(prop_uplisted ~ 1 + (1 | Order), data = con_status_interacts_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

# use anova to compare
anova(con_status_uplisted_glmm_EDGE, con_status_uplisted_glmm_EDGE_null)

### create tabular format or results
tab_model(file = "con_status_uplisted_glmm_EDGE.doc",
  con_status_uplisted_glmm_ED, con_status_uplisted_glmm_EDGE,
  show.est = TRUE, 
  string.est = "Estimate", 
  transform = NULL, 
  digits = 4,
  CSS = list(
    css.depvarhead = 'color: black;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: black;'
  )
)

webshot("con_status_uplisted_glmm_EDGE.html", "con_status_uplisted_glmm_EDGE.png")

#extract effects for conservation uplistings for EDGE scores
# do same for ED
uplisting_df_EDGE <- effects::effect(term = c("median_EDGE"), mod = con_status_uplisted_glmm_EDGE) %>%
  # turn it into a dataframe
  as.data.frame()

uplisting_df_EDGE$Predictor <- "Median EDGE"
names(uplisting_df_EDGE)[1] <- "proportion"

tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_prop_EDGE_uplisting.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot for EDGE scores and conservation up-listings
p <- ggplot(uplisting_df_EDGE) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower, ymax = upper, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Median EDGE" ,y = "Proportion of Red List up-listings")

# change colour pallete to orange
grid.arrange((p + scale_colour_brewer(palette = "Set1") + theme_Publication()),nrow=1)

dev.off()

```

## Question (iii) How does Is the response of population status/family associated with the concentration of anthropogenic threats for priority lineages (REPHRASE question in the morning)

```{r population trend analysis, echo=TRUE}

#subset threat data
pop_status_df <- mammals_pd_merged[,c(1:2,9,43,45,39,49)]

# Remove all NA's from columns
pop_status_df <- pop_status_df[complete.cases(pop_status_df),]

# add column to add observational level occurences
pop_status_df$observation <- 1:nrow(pop_status_df)

pop_status_df$observation <- as.numeric(pop_status_df$observation)

# visulising distibutions of data
par(mfrow=c(4,2))
for(i in 3:7) {
    hist(pop_status_df[,i], main=names(pop_status_df)[i])
}

# creating a correlation matrix - no correlations between variables. Can therefore dismiss the assumption of non-independence
correlations <- cor(pop_status_df[,3:14])
corrplot(correlations, method="circle")

# glm in which EDGE or ED scores is a responds to proportion of threatened species interacts with threats predict popualtion decline
# perform a glm fit
pop_status_dec_glmm_ED <- glmer(Prop_dec ~ assessed_threat_prop*median_ED + (1 | Order), data = pop_status_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

# test for overdispersion
performance::check_overdispersion(pop_status_dec_glmm_ED)

# get tabular of model
tab_model(pop_status_dec_glmm_ED)

# get explanatory power
StatisticalModels::R2GLMER(pop_status_dec_glmm_ED)
MuMIn::r.squaredGLMM(pop_status_dec_glmm_ED)

# check model performance
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/pop_status_dec_glmm_ED_crit.tiff", units = "in", width = 8, height = 10, res = 300)

check_model(pop_status_dec_glmm_ED)

dev.off()


# create null model
pop_status_dec_glmm_ED_null <- glmer(Prop_dec ~ 1 + (1 | Order), data = pop_status_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

# comapre random effect - keep random effect
anova(pop_status_dec_glmm_ED, pop_status_dec_glmm_ED_null)

### Create tabular format of results
### create tabular format or results
tab_model(file = "pop_dec_glmm_ED.html",
  pop_status_dec_glmm_ED,
  show.est = TRUE, 
  string.est = "Estimate", 
  transform = NULL, 
  digits = 4,
  CSS = list(
    css.depvarhead = 'color: black;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: black;'
  )
)

webshot("pop_status_dec_glmm_ED.html", "pop_status_dec_glmm_ED.png")

# Predict conservation up-listings for ED:threatened species interaction
# pull out the significant effects of the model
EDGE_pop_dec_df_threat <- effects::effect(term = c("assessed_threat_prop"), mod = pop_status_dec_glmm_ED) %>%
  # turn it into a dataframe
  as.data.frame()

EDGE_pop_dec_df_threat$Predictor <- "Threatened species"
names(EDGE_pop_dec_df_threat)[1] <- "proportion"

# Predict conservation up-listings for ED:threatened species interaction
# pull out the significant effects of the model
ED_pop_dec_df <- effects::effect(term = c("median_ED"), mod = pop_status_dec_glmm_ED) %>%
  # turn it into a dataframe
  as.data.frame()

ED_pop_dec_df$Predictor <- "Median ED"
names(ED_pop_dec_df)[1] <- "proportion"



tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_prop_dec_threat.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot
p <- ggplot(EDGE_pop_dec_df_threat) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower, ymax = upper, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Proportion of family threatened" ,y = "Proportion of population decreasing")

# change colour pallete to orange
grid.arrange((p + scale_colour_brewer(palette = "Spectral", direction = -1) + theme_Publication()),nrow=1)

dev.off()

# plot predictions for ED and proportion of decreases
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/ED_glmm_prop_dec.tiff", units = "in", width = 7, height = 5, res = 300)

# set up a ggplot
p <- ggplot(ED_pop_dec_df) + 
  # plot the average slope (fit) for each significant effect
  # I've used automatic colouring for each effect
  geom_line(aes(x = proportion, y = fit, group = Predictor, color = Predictor), alpha = 5, size = 1) +
  # add shading for the confidence intervals for each slope
  geom_ribbon(aes(x = proportion, ymin = lower, ymax = upper, group = Predictor), alpha = 0.1) +
  # add the x and y label
  labs(x = "Median ED" ,y = "Proportion of population decreasing")

# change colour pallete to orange
grid.arrange((p + scale_colour_brewer(palette = "Set1") + theme_Publication()),nrow=1)

dev.off()

#################################################
### Look at response of EDGE scores on decreasing population 
pop_status_dec_glmm_EDGE <- glmer(Prop_dec ~ median_EDGE + (1 | Order), data = pop_status_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

# Create null model
pop_status_dec_glmm_EDGE_null <- glmer(Prop_dec ~ 1 + (1 | Order), data = pop_status_df, family = binomial(link = "logit"), control = glmerControl(optimizer = "bobyqa"))

# compare null model
anova(pop_status_dec_glmm_EDGE, pop_status_dec_glmm_ED_null)

# look at model summary
summary(pop_status_dec_glmm_EDGE)

# test for overdispersion
performance::check_overdispersion(pop_status_dec_glmm_EDGE)

# get tabular of model
tab_model(pop_status_dec_glmm_EDGE,
          show.est = TRUE,
          string.est = "Estiamte",
          transform = NULL)

# get explanatory power
StatisticalModels::R2GLMER(pop_status_dec_glmm_EDGE)
MuMIn::r.squaredGLMM(pop_status_dec_glmm_ED)

# check model performance
tiff("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Results/Conservation status/pop_status_dec_glmm_EDGE_crit.tiff", units = "in", width = 8, height = 10, res = 300)

performance::check_overdispersion(pop_status_dec_glmm_EDGE)

dev.off()

### create tabular format or results
tab_model(file = "pop_status_dec_glmm_ALL.doc",
  pop_status_dec_glmm_ED, pop_status_dec_glmm_EDGE,
  show.est = TRUE,
  string.est = "Estiamte",
  transform = NULL,
  digits = 4,
  CSS = list(
    css.depvarhead = 'color: black;',
    css.centeralign = 'text-align: left;', 
    css.firsttablecol = 'font-weight: bold;', 
    css.summary = 'color: black;'
  )
)

webshot("pop_status_dec_glmm_ALL.html", "pop_status_dec_glmm_ALL.png")

# extract effects for EDGE score
EDGE_pop_dec_df <- effects::effect(term = c("median_EDGE"), mod = pop_status_dec_glmm_EDGE) %>%
  # turn it into a dataframe
  as.data.frame()

EDGE_pop_dec_df$Predictor <- "Median EDGE"
names(EDGE_pop_dec_df)[1] <- "proportion"

```


# Identifying EDGE summary statistics and graphs
```{r priority lineages, echo=TRUE}

# number of EDGE lineages - 5
sum(EDGE_priority_lineages$no_DD_NE_prop > 0.5)

EDGE_priority_lineages <- EDGE_priority_lineages[c(EDGE_priority_lineages$no.over.median > 94),]

# Get lengths of each priority lineage list 
sum(lengths(regmatches(EDGE_priority_lineages$priority_type, gregexpr("Watchlist lineage", EDGE_priority_lineages$priority_type))))

priority_sub <- EDGE_priority_lineages[c(EDGE_priority_lineages$priority_type == "Priority lineage"),]

# Calculate number of occurences which are greater than 0 to sum threat data
priority_sub$threat_sum <- apply(priority_sub[,c("prop_1_1","prop_2_2","prop_2_3","prop_5_1","prop_5_3","prop_8_1")], 1, function(x) length(which(x > 0)))

# find the percentage of priority EDGE lineages decreasing
EDGE_prioirty <- subset(mammals_pd_merged, priority_type == "Priority lineage")

EDGE_priority_lineages <- subset(mammals_pd_merged, EDGE_priority == "Yes")

# sum number of lineages that are decreasing = 20
(sum(EDGE_prioirty$Prop_dec == 1, na.rm = TRUE)/23)*100




```
