---
title: "PHASE_THREE_STATISTICAL_ANALYSIS"
author: "Chris Low"
date: "01/09/2021"
output: html_document
---

```{r setup, include=FALSE}
library(performance) # check model outputs
library(lme4) # statistical modelling for logistic regressions and mixed-effects modelling
library(corrplot) # creating a correlation matrix

```

## Loading in final dataset in r

```{r load data, echo=TRUE}

# Just save priority EDGE lineage list and PD_mammal_merged list as essential for analysis: mammals_pd_merged, EDGE_priority_lineages (final dataset so far!!!)
load("C:/Users/Student/Zoological Society of London/EDGEofExistence - Chris Low - EDGE lineages/Project II/Data and analysis/EDGE_families_and_scores/EDGE_LINEAGES_FOR_ANALYSIS.RData")

# load dataout datasets with both Upham and Gumbs datasets

```

## Testing the difference between

You can also embed plots, for example:

```{r pressure, echo=FALSE}

cor.test(mammals_pd_merged$Median.exPD, mammals_pd_merged$Median.ePD.UPHAM, method = "spearman")

```
## Question (i) is the response of ePD loss a good predictor of extinction risk?

## Question (ii) Is the conservation status of priority lineages in a greater state of decline than non-priority lineages

```{r explore data, echo=TRUE}

names(mammals_pd_merged)

head(mammals_pd_merged)

summary(mammals_pd_merged)

# visulising distibutions of data
par(mfrow=c(2,46))
for(i in 2:46) {
    hist(mammals_pd_merged[,i], main=names(mammals_pd_merged)[i])
}

# creating a correlation matrix
correlations <- cor(mammals_pd_merged[,2:46])
corrplot(correlations, method="circle")

# convert prioirty values to 1's and 0's
mammals_pd_merged$EDGE_priority[mammals_pd_merged$EDGE_priority == "No"] <- 0
mammals_pd_merged$EDGE_priority[mammals_pd_merged$EDGE_priority == "Yes"] <- 1

# change to binary operator
mammals_pd_merged$EDGE_priority <- as.numeric(mammals_pd_merged$EDGE_priority)

# perform a glm fit
priority_lineage_glm <- glm(EDGE_priority ~ Prop_dec + Prop_inc + Prop_stab + prop_downlisted + prop_uplisted + prop_stable, family = binomial, data = mammals_pd_merged)

summary(priority_lineage_glm)

```

## Question (iii) Is the response of population status/family associated with the concentration of anthropogenic threats for priority lineages
